---
title: "fitting gams with no forcing through origin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we fit gams with no forcing through origin

libraries
`````{r}
library(knitr)
library(mgcv)
library(dplyr)
library(reshape2)
library(ggplot2)
library(tidymv)
```````
data
``````{r}
a<-read.csv("org_and_cult_haldrates_before_gams.csv")
``````
get how many intervals there are for each series
``````{r}
a<-a%>%
group_by(population, variable)%>%
mutate(max_intervals=max(interval))
```````
get a unique designator for each series
`````{r}
a$popvar<-paste(a$population, a$variable, sep="_")
``````
make a function to fit a gam
``````{r}
onepopvar<-function(onepopvar){
model<- gam(abs.hald.num ~ s(interval, bs = 'cr'), data=onepopvar, family=gaussian)
interval<-1:100
pred<-as.data.frame(interval)
p1<-predict.gam(model,pred,se=TRUE)
pred$fit<-p1$fit
pred$se<-p1$se
return(pred)
}
```````
run the model over all popvar
``````{r}
r1<-a %>% group_by(popvar) 
%>% do(onepopvar(.))
r2<-as.data.frame(r1)
````````

`````{r}
a3<-unique(a[c("popvar", "max_intervals")])
r3<-merge(r2, a3, by="popvar")
r3<-subset(r3, interval<=max_intervals)
```````
check the fit of a single popvar
`````{r}
popvars<-unique(r3$popvar)
popvars
``````

`````{r}
a_check<-subset(a, popvar=="Cepaea_nemoralis_banded")
r_check<-subset(r3, popvar=="Cepaea_nemoralis_banded")
a_check_means<-a_check%>%
group_by(interval)%>%
summarize(mean.abs.hald.num=mean(abs.hald.num))
```````
plot abs.hald.num
`````{r}
ggplot()+
geom_point(data=a_check, aes(x=interval, y=abs.hald.num), size=1, colour="grey")+
geom_point(data=a_check_means, aes(x=interval, y=mean.abs.hald.num), size=1, colour="black")+
geom_line(data=r_check, aes(x=interval, y=fit), size=1, colour="red")+
theme_classic()
```````
### check h_1

``````{r}
a_check_means<-a%>%
group_by(population, popvar, interval)%>%
summarize(mean.abs.hald.num=mean(abs.hald.num))%>%
filter(interval==1)
`````````

``````{r}
r_check<-r3%>%
filter(interval==1)
r_check$interval<-NULL
````````

``````{t}
r4<-merge(a_check_means, r_check, by="popvar")
````

`````{r}
x<-0:15
y<-0:15
equal<-as.data.frame(cbind(x,y))
```````

```{r}
ggplot()+
geom_point(data=r4, aes(x=mean.abs.hald.num, y=fit, colour=as.factor(population)), size=1)+
geom_line(data=equal, aes(x=x, y=y), size=0.5, colour="red")+
theme_classic()+
theme(aspect.ratio=1)
```````

```{r}
ggplot()+
geom_point(data=r4, aes(x=log(mean.abs.hald.num), y=log(fit), colour=as.factor(population)), size=1)+
geom_line(data=equal, aes(x=log(x), y=log(y)), size=0.5, colour="red")+
theme_classic()+
theme(aspect.ratio=1)
```````
why is G. aculeatus so low? 
``````{r}
check<-subset(r4, population=="Gasterosteus_aculeatus")
````````
### check h-25

``````{r}
a_check_means<-a%>%
group_by(population, popvar, interval)%>%
summarize(mean.abs.hald.num=mean(abs.hald.num))%>%
filter(interval==25)
`````````

``````{r}
r_check<-subset(r3, interval==25)
r_check$interval<-NULL
````````

``````{t}
r_check<-as.data.frame(r_check)
a_check_means<-as.data.frame(a_check_means)
r5<-merge(a_check_means, r_check, by.x="popvar", by.y="popvar")
````

`````{r}
x<-0:40
y<-0:40
equal<-as.data.frame(cbind(x,y))
```````

```{r}
ggplot()+
geom_point(data=r5, aes(x=mean.abs.hald.num, y=fit, colour=as.factor(population)), size=1)+
geom_line(data=equal, aes(x=x, y=y), size=0.5, colour="red")+
theme_classic()+
theme(aspect.ratio=1)
```````

```{r}
ggplot()+
geom_point(data=r5, aes(x=log(mean.abs.hald.num), y=log(fit), colour=as.factor(population)), size=1)+
geom_line(data=equal, aes(x=log(x), y=log(y)), size=0.5, colour="red")+
theme_classic()+
theme(aspect.ratio=1)
```````
So, they're all good except for Jadera which has a low sample size and the values are all over the place.

look at the haldane rates themselves
``````{r}
r3$haldane<-r3$fit/r3$interval
a2<-unique((a[c("population", "popvar")]))
r10<-merge(a2, r3, by="popvar")
r10<- r10[order(r10$popvar, r10$interval),] 
```````

```{r}
ggplot()+
geom_line(data=r10, aes(x=interval, y=haldane, colour=as.factor(population), group=popvar), size=0.5)+
theme_classic()+
theme(aspect.ratio=1)
```````
#### checking individual values
Check individual values
``````{r}
check<-subset(r10, population=="Gasterosteus_aculeatus" & interval==1)
check2<-a%>% 
group_by(population, popvar, interval)%>%
summarize(hald.rate=mean(hald.rate))%>%
filter(population=="Gasterosteus_aculeatus" & interval==1)
check2<-as.data.frame(check2)
```````
So they all seem to be OK except for Gasterosteus aculeatus which seems low. 
Let's see if we can get a better fit by increasing the number of k
#### re-running Gasterosteus with higher k

````{r}
ga<-subset(a, population=="Gasterosteus_aculeatus")
```````
make a function to fit a gam
``````{r}
onepopvar<-function(onepopvar){
model<- gam(abs.hald.num ~ s(interval, bs = 'cr', k=20), data=onepopvar, family=gaussian)
interval<-1:100
pred<-as.data.frame(interval)
p1<-predict.gam(model,pred,se=TRUE)
pred$fit<-p1$fit
pred$se<-p1$se
return(pred)
}
```````
run the model over all popvar
``````{r}
ga2<-ga %>% group_by(popvar) %>% do(onepopvar(.))
ga2<-as.data.frame(ga2)
ga2$haldane<-ga2$fit/ga2$interval
head(ga2)
````````
that doesn't help.  I think we are just going to have to live with that. 

#### get CIs of fits

``````{r}
r3$LCI<-r3$fit-r3$se*1.96
r3$UCI<-r3$fit+r3$se*1.96
r3$haldane_LCI<-r3$LCI/r3$interval
r3$haldane_UCI<-r3$UCI/r3$interval
r3$haldane_CI<-(r3$haldane_UCI-r3$haldane_LCI)/2
````````

``````{r}
a5<-unique(a[c("population", "variable", "popvar")])
`````````

``````{r}
r5<-merge(r3, a5, by="popvar")
write.csv(r5, "all_cult_organic_haldane_gams_july_2019.csv")
``````


